---
title: 'Validation of Fuzzy and Possibilistic Clustering'
author: 'Zeynel Cebeci'
date: 'December 13, 2017'
output:
  prettydoc::html_pretty:
    theme: tactile
    highlight: github
    df_print: paged
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Validation of Fuzzy and Possibilistic Clustering Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
<style>
body{
text-align: justify}
</style>

# Data set and Packages

## Install and load the package fcvalid
In order to validate the results of fuzzy and possibilistic cluster analysis, this vignette is designed to demonstrate the use of some functions in the package '`fcvalid`'. You can download the recent version of the package from CRAN with the following command:

```{r, eval=FALSE, message=FALSE, warning=FALSE}
if(!require(prettydoc) {install.packages('prettydoc')}

if(!require(devtools) {install.packages('devtools'); library(devtools)}
 devtools::install_github("zcebeci/fcvalid",build_vignettes = TRUE)
```

If you have already installed '`fcvalid`', you can load it into R working environment by using the following command:

```{r, echo=TRUE, message=FALSE, warning=FALSE}
library(fcvalid)
```

## Load the data set 
We demonstrate clusterig validation using the package '`fcvalid`' on Iris data set (Anderson, 1935), a well-known real data set consisting of four features (`Sepal.Length`, `Sepal.Width`, `Petal.Length` and `Petal.Width`) with natural classes of three Iris species on the last column of data set. This four-dimensional data set contains totally 150 data objects, 50 in each class.

```{r echo=TRUE, message=FALSE, warning=FALSE, cols.print=5, rows.print=10}
data(iris)
head(iris)
tail(iris)
x <- iris[,-5]
```
The following R command displays the scatter plots between the feature pairs. 

```{r fig.width=7, fig.height=6}
pairs(x, col=iris[,5])
```

# Partitioning Cluster Analysis
In this vignette we demonstrate the use of internal indexes with two partitioning clustering algorithms in the package 'ppclust' (Cebeci et al, 2017). Thus, we first load this package into R working environment.


```{r, echo=TRUE, message=FALSE, warning=FALSE}
library(ppclust)
```

## Fuzzy C-Means Clustering

As the representative of the fuzzy clustering algorithms for demonstrating cluster validation we'll use the basic Fuzzy C-Means (FCM) clustering algorithm proposed by Bezdek (1974). We run `fcm` function of the `ppclust` package for three clusters as follows:

```{r, echo=TRUE, message=FALSE, warning=FALSE}
cres <- fcm(x, centers=3, m=2)
```

In order to summarize and plot the clustering results, the functions '`summary`' and '`plotcluster`' of the package '`ppclust`' can be used as follows:

```{r, echo=TRUE, message=FALSE, warning=FALSE}
summary(cres)
plotcluster(cres, trans=TRUE)
```

## Clustering Validity
In order to validate the clustering result obtained in the previous subsection by using FCM, any of the cluster validity index of the package '`fcvalid`. The following two commands demonstrate the use of the Partition Coeffient and Xie-Beni indexes.

```{r, echo=TRUE, message=FALSE, warning=FALSE}
pc(cres)
xb(cres)
```
Normally, however the users may evaluate the clustering quality with a couple of the internal indexes individually as shown above, the validity measures of all of the indexes may also be obtained alltogether. For this purpose, the function `allindexes` of the package '`fcvalid`' is used as follows:

```{r, echo=TRUE, message=FALSE, warning=FALSE}
allindexes(cres)
```

## Possibilistic Fuzzy C-Means Clustering
In this subsection, Possibilistic Fuzzy C-Means (PFCM) (Pal et al, 2005) is run to demonstrate the extended and generalized versions of the internal indexes. 

```{r, echo=TRUE, message=FALSE, warning=FALSE}
cres2 <- pfcm(x=x, centers=3, m=2, eta=2)
```

In order to summarize and plot the clustering results, the functions '`summary`' and '`plotcluster`' of the package '`ppclust`' can be used as follows:

```{r, echo=TRUE, message=FALSE, warning=FALSE}
summary(cres2)
plotcluster(cres2, trans=TRUE)
```

PFCM algorithm produces both the fuzzy and possibilistic partitions of data sets. For this reason the internal fuzzy validation indexes cannot be directly applied to evaluate clustering results of PFCM as well as other fuzzy and possibilistic clustering algorithms. In such cases, the generalized and extended versions of the fuzzy validity indexes (Cebeci et al, 2017) can be used for validation of the possibilistic clustering results. In the following example, the generalized index values of a PFCM run are shown for the indexes of Tang-Sun & Sun (TSS) and Fukuyama-Sugeno (FS).

```{r, echo=TRUE, message=FALSE, warning=FALSE}
fs(cres2, tidx="g")
tss(cres2, tidx="g")
```
Alternatively, the extended index values can be computed to evaluate the fuzzy and possibilistic clustering results. An extended index value is obtained by summation of fuzzy and possibilistic membership degrees for the algorithms producing both types of partitions. The example below demonstrates how to handle the extended values for XB and PBM indexes:

```{r, echo=TRUE, message=FALSE, warning=FALSE}
xb(cres2, tidx="e")
pbm(cres2, tidx="e")
```
In order to compute the generalized index values from all internal indexes, the function `allindexes` can be run as follows:

```{r, echo=TRUE, message=FALSE, warning=FALSE}
allindexes(cres2, tidx="g")
```

# Finding Optimal Clustering 
In order to decide to the best clustering result or to find an optimal value of number of clusters in data sets, cluster analysis should be repeated for a range of number of clusters. In the code chunk below, FCM algorithm is run for five levels of `k` (from 2 to 6). The lower (`k1`) and upper (`k2`) values can be changed for testing the different ranges of `k`.

```{r, echo=TRUE, message=FALSE, warning=FALSE}
k1 <- 2
k2 <- 6
indnames <- c("pc","mpc","pe","xb","kwon", "tss", "cl", "fs", 
              "pbm","si","sc","awcd", "fhv", "apd", "cs")
indvals <- matrix(ncol=length(indnames), nrow=k2-k1+1)
i <- 1
for(k in k1:k2){
  cres <- fcm(x=x, centers=k, nstart=3)
  indvals[i,1] <- pc(cres)
  indvals[i,2] <- mpc(cres)
  indvals[i,3] <- pe(cres)
  indvals[i,4] <- xb(cres)
  indvals[i,5] <- kwon(cres)
  indvals[i,6] <- tss(cres)
  indvals[i,7] <- cl(cres)
  indvals[i,8] <- fs(cres)
  indvals[i,9] <- pbm(cres)
  indvals[i,10] <- si(cres)$sif
  indvals[i,11] <- sc(cres)
  indvals[i,12] <- awcd(cres)
  indvals[i,13] <- fhv(cres)
  indvals[i,14] <- apd(cres)
  indvals[i,15] <- cs(cres)
  i <- i+1
}
colnames(indvals) <- indnames
rownames(indvals) <- paste0("k=",k1:k2) 
print(indvals)
```
The computed index values can be visually inspected by using barplots, plots or other kinds of graphical tools.

```{r fig.width=9, fig.height=11}
par(mfrow=c(5,3))
for(i in 1:length(indnames))
  barplot(indvals[,i], col="dodgerblue", main=indnames[i])
```

```{r fig.width=9, fig.height=11}
par(mfrow=c(5,3))
for(i in 1:length(indnames)){
  plot(0,0, type = "n",  
	cex.lab=0.8, cex.axis=0.8, cex.main=0.8, cex.sub=0.8, 
	xlim = c(1, nrow(indvals)), ylim = c(min(indvals[,i]),max(indvals[,i])), 
	xaxt='n', xlab="number of clusters", ylab="index value",
	main=indnames[i], sub=" ")
  axis(side=1, at=seq(1, nrow(indvals), by=1), 
    labels=paste0("k=",k1:k2), col.axis="black", las=1)
  lines(indvals[,i], type="b", col="blue", lty=1, lwd=1)
}
```


# References
Anderson, E. (1935). The irises of the GASPE peninsula, in *Bull. Amer. Iris Soc.*, 59: 2-5.

Bezdek, J.C. (1974). Cluster validity with fuzzy sets. *J Cybernetics*, 3(3):58-72. https://doi.org/10.1080/01969727308546047

Pal, N. R., Pal, K. & Bezdek, J. C. (2005). A possibilistic fuzzy c-means clustering algorithm. *IEEE Trans. Fuzzy Systems*, 13 (4): 517-530. https://doi.org/10.1109/TFUZZ.2004.840099

Cebeci, Z., Kavlak, A. T., & Yildiz, F. (2017, September). Validation of fuzzy and possibilistic clustering results. In 2017 International Artificial Intelligence and Data Processing Symposium (IDAP) (pp. 1-7). IEEE.

Cebeci, Z., Yildiz, F., Kavlak, A.T., Cebeci, C. & Onder, H. (2017). ppclust: Probabilistic and Possibilistic Cluster Analysis. R package  version 0.1.0, URL https://CRAN.R-project.org/package=ppclust
